---
title: "week4"
author: "Ilir_Sheraj"
date: "8/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Gapminder Data

Data visualization can be used to dispel common myths and educate the public and contradict sensationalist or outdated claims and stories. We will use real data to answer the following questions about world health and economics:

1 - Is it still fair to consider the world as divided into the West and the developing world?

2 - Has income inequality across countries worsened over the last 40 years?

```{r}
library(dslabs)
library(dplyr)
data("gapminder")
str(gapminder)
```

```{r}
dim(gapminder)
```

```{r}
gapminder %>% filter(year == 2015 & country %in% c("Sri Lanka", "Turkey")) %>% 
  select(country, infant_mortality)
```

# Life Expectancy and Fertility Rates

```{r}
library(ggplot2)
gapminder %>% filter(year == 1962) %>% 
  ggplot(aes(fertility, life_expectancy)) +
  geom_point() +
  ds_theme_set()
```

```{r}
gapminder %>% filter(year == 1962) %>% 
  ggplot(aes(fertility, life_expectancy, color = continent)) +
  geom_point() +
  ds_theme_set()
```

# Faceting

```{r}
gapminder %>% filter(year == 2012) %>% 
  ggplot(aes(fertility, life_expectancy, color = continent)) +
  geom_point() +
  ds_theme_set()
```

```{r}
gapminder %>% filter(year %in% c(1962,2012)) %>% 
  ggplot(aes(fertility, life_expectancy, color = continent)) +
  geom_point() +
  ds_theme_set() +
  facet_grid(continent ~ year)
```

```{r}
gapminder %>% filter(year %in% c(1962,2012)) %>% 
  ggplot(aes(fertility, life_expectancy, color = continent)) +
  geom_point() +
  ds_theme_set() +
  facet_grid(.~year)
```

```{r}
years <- c(1962, 1980, 1990, 2000, 2012)
continents <- c("Europe", "Asia")
gapminder %>%
    filter(year %in% years & continent %in% continents) %>%
    ggplot(aes(fertility, life_expectancy, col = continent)) +
    geom_point() +
    facet_wrap(~year)
```

# Time Series Plots

```{r}
gapminder %>%
    filter(country == "United States") %>%
    ggplot(aes(year, fertility)) +
    geom_point()
```

```{r}
gapminder %>%
    filter(country == "United States") %>%
    ggplot(aes(year, fertility)) +
    geom_line()
```

```{r}
gapminder %>%
    filter(country %in% c("South Korea", "Germany")) %>%
    ggplot(aes(year, fertility, col = country)) +
    geom_line()
```

```{r}
countries <- c("South Korea", "Germany")
labels <- data.frame(country = countries, x = c(1975, 1965), y = c(60, 72))
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, life_expectancy, col = country)) +
    geom_line() +
    geom_text(data = labels, aes(x, y, label = country), size = 5) +
    theme(legend.position = "none")
```

# Transformations

```{r}
gapminder <- gapminder %>% 
  mutate(dollars_per_day = gdp/population/365)
past_year <- 1970
gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>% 
  ggplot(aes(dollars_per_day)) +
  geom_histogram(binwidth = 1, color = "black")
```

```{r}
gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>% 
  ggplot(aes(log2(dollars_per_day))) +
  geom_histogram(binwidth = 1, color = "black")
```

```{r}
gapminder %>% 
  filter(year == past_year & !is.na(population)) %>% 
  ggplot(aes(log10(population))) +
  geom_histogram(binwidth = 0.5, color = "black")
```


```{r}
gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2")
```

# Stratify and Boxplot

```{r}
length(levels(gapminder$region))
```

```{r}
p <- gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(region, dollars_per_day))
p + geom_boxplot()
```

```{r}
p + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
fac <- factor(c("Asia", "Asia", "West", "West", "West"))
levels(fac)
```

```{r}
value <- c(10, 11, 12, 6, 4)
fac <- reorder(fac, value, FUN = mean)
levels(fac)
```

```{r}
p <- gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%    # reorder
    ggplot(aes(region, dollars_per_day, fill = continent)) +    # color by continent
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("")
p
```

```{r}
gapminder %>%
    filter(year == 2011 & !is.na(gdp)) %>%
    mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%    # reorder
    ggplot(aes(region, dollars_per_day, fill = continent)) +    # color by continent
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("")
```

```{r}
p + scale_y_continuous(trans = "log2")
```

```{r}
gapminder %>%
    filter(year == 2011 & !is.na(gdp)) %>%
    mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%    # reorder
    ggplot(aes(region, dollars_per_day, fill = continent)) +    # color by continent
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("") + scale_y_continuous(trans = "log2")
```

```{r}
p + scale_y_continuous(trans = "log2") + geom_point(show.legend = FALSE)
```

# Comparing Distributions

```{r}
gapminder <- gapminder %>%
    mutate(dollars_per_day = gdp/population/365)
past_year <- 1970
# define Western countries
west <- c("Western Europe", "Northern Europe", "Southern Europe", "Northern America", "Australia and New Zealand")
gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2") +
    facet_grid(. ~ group)
```

```{r}
past_year <- 1970
present_year <- 2010
gapminder %>%
    filter(year %in% c(past_year, present_year) & !is.na(gdp)) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2") +
    facet_grid(year ~ group)
```

```{r}
country_list_1 <- gapminder %>%
    filter(year == past_year & !is.na(dollars_per_day)) %>% .$country
country_list_2 <- gapminder %>%
    filter(year == present_year & !is.na(dollars_per_day)) %>% .$country
country_list <- intersect(country_list_1, country_list_2)
gapminder %>% 
    filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2") +
    facet_grid(year ~ group)
```

```{r}
p <- gapminder %>%
    filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
    mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%
    ggplot(aes(region, dollars_per_day)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("") + scale_y_continuous(trans = "log2")
```

```{r}
p + geom_boxplot(aes(region, dollars_per_day, fill = continent)) +
    facet_grid(year ~ .)
```

```{r}
p + geom_boxplot(aes(region, dollars_per_day, fill = factor(year)))
```

# Density Plots

```{r}
#gapminder %>%
#    filter(year == past_year & country %in% country_list) %>%
#    mutate(group = ifelse(region %in% west, "West", "Developing")) %>% group_by(group) %>%
#    summarize(n = n()) %>% knitr::kable()
```

```{r}
p <- gapminder %>%
    filter(year == past_year & country %in% country_list) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day, y = ..count.., fill = group)) +
    scale_x_continuous(trans = "log2")
p + geom_density(alpha = 0.2)
```

```{r}
p <- gapminder %>%
  filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
  mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
  ggplot(aes(dollars_per_day, y = ..count.., fill = group)) +
  scale_x_continuous(trans = "log2")
# stacked density plot
p + geom_density(alpha = 0.2) +
    facet_grid(year ~ .)
```

```{r}
p <- gapminder %>%
  filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
  mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
  ggplot(aes(dollars_per_day, y = ..count.., fill = group)) +
  scale_x_continuous(trans = "log2")
# stacked density plot
p + geom_density(alpha = 0.2, bw = 0.75) +
    facet_grid(year ~ .)
```

## case_when()

```{r}
gapminder <- gapminder %>%
    mutate(group = case_when(
            .$region %in% west ~ "West",
            .$region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
            .$region %in% c("Caribbean", "Central America", "South America") ~ "Latin America",
            .$continent == "Africa" & .$region != "Northern Africa" ~ "Sub-Saharan Africa",
            TRUE ~ "Others"))
gapminder <- gapminder %>%
    mutate(group = factor(group, levels = c("Others", "Latin America", "East Asia", "Sub-Saharan Africa", "West")))

p <- gapminder %>%
  filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
    ggplot(aes(dollars_per_day, fill = group)) +
    scale_x_continuous(trans = "log2")
p + geom_density(alpha = 0.2, bw = 0.75) +
    facet_grid(year ~ .)
```

```{r}
p + geom_density(alpha = 0.2, bw = 0.75, position = "stack") +
    facet_grid(year ~ .)
```

```{r}
gapminder %>%
    filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
    group_by(year) %>%
    mutate(weight = population/sum(population*2)) %>%
    ungroup() %>%
    ggplot(aes(dollars_per_day, fill = group, weight = weight)) +
    scale_x_continuous(trans = "log2") +
    geom_density(alpha = 0.2, bw = 0.75, position = "stack") + facet_grid(year ~ .)
```

# Ecological Fallacy

```{r}
gapminder <- gapminder %>%
    mutate(group = case_when(
        .$region %in% west ~ "The West",
        .$region %in% "Northern Africa" ~ "Northern Africa",
        .$region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
        .$region == "Southern Asia" ~ "Southern Asia",
        .$region %in% c("Central America", "South America", "Caribbean") ~ "Latin America",
        .$continent == "Africa" & .$region != "Northern Africa" ~ "Sub-Saharan Africa",
        .$region %in% c("Melanesia", "Micronesia", "Polynesia") ~ "Pacific Islands"))
surv_income <- gapminder %>%
    filter(year %in% present_year & !is.na(gdp) & !is.na(infant_mortality) & !is.na(group)) %>%
    group_by(group) %>%
    summarize(income = sum(gdp)/sum(population)/365,
                        infant_survival_rate = 1 - sum(infant_mortality/1000*population)/sum(population))
surv_income %>% arrange(income)
```

```{r}
surv_income %>% ggplot(aes(income, infant_survival_rate, label = group, color = group)) +
    scale_x_continuous(trans = "log2", limit = c(0.25, 150)) +
    scale_y_continuous(trans = "logit", limit = c(0.875, .9981),
                                       breaks = c(.85, .90, .95, .99, .995, .998)) +
    geom_label(size = 3, show.legend = FALSE) 
```

logit transformation: f(p) = log(p/(1-p)). This gives the odds of children survival. 

*Ecological Fallacy*

# Assessment

Using ggplot and the points layer, create a scatter plot of life expectancy versus fertility for the African continent in 2012.

```{r}
gapminder %>% filter(year == 2012 & continent == "Africa") %>%
  ggplot(aes(fertility, life_expectancy)) +
  geom_point()
```

Remake the plot from the previous exercises but this time use color to distinguish the different regions of Africa to see if this explains the clusters. Remember that you can explore the gapminder data to see how the regions of Africa are labeled in the data frame! Use color rather than col inside your ggplot call - while these two forms are equivalent in R, the grader specifically looks for color.

```{r}
gapminder %>% filter(year == 2012 & continent == "Africa") %>%
  ggplot(aes(fertility, life_expectancy, color = region)) +
  geom_point()
```

Create a table showing the country and region for the African countries (use select) that in 2012 had fertility rates of 3 or less and life expectancies of at least 70. Assign your result to a data frame called df.

```{r}
df <- gapminder %>% filter(year == 2012 & continent == "Africa" & life_expectancy >= 70 & fertility <= 3) %>%
  select(country, region)
df
```

The Vietnam War lasted from 1955 to 1975. Do the data support war having a negative effect on life expectancy? We will create a time series plot that covers the period from 1960 to 2010 of life expectancy for Vietnam and the United States, using color to distinguish the two countries. In this start we start the analysis by generating a table. Use filter to create a table with data for the years from 1960 to 2010 in Vietnam and the United States. Save the table in an object called tab.

```{r}
tab <- gapminder %>% filter(country %in% c("Vietnam", "United States") & year <= 2010)
```

Use geom_line to plot life expectancy vs year for Vietnam and the United States and save the plot as p. The data table is stored in tab. Use color to distinguish the two countries.

```{r}
p <- tab %>% ggplot(aes(year, life_expectancy, color = country)) +
  geom_line()
p
```

Cambodia was also involved in this conflict and, after the war, Pol Pot and his communist Khmer Rouge took control and ruled Cambodia from 1975 to 1979. He is considered one of the most brutal dictators in history. Do the data support this claim? Use a single line of code to create a time series plot from 1960 to 2010 of life expectancy vs year for Cambodia.

```{r}
gapminder %>% filter(country == "Cambodia" & year <= 2010) %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_line()
```

Now we are going to calculate and plot dollars per day for African countries in 2010 using GDP data. In the first part of this analysis, we will create the dollars per day variable. Use mutate to create a dollars_per_day variable, which is defined as gdp/population/365. Create the dollars_per_day variable for African countries for the year 2010. Remove any NA values. Save the mutated dataset as daydollars.

```{r}
daydollars <- gapminder %>% 
  mutate(dollars_per_day = gdp/population/365) %>% 
  filter(continent == "Africa" & year == 2010 & !is.na(dollars_per_day))
```

Create a smooth density plot of dollars per day from daydollars. Use scale_x_continuous to change the x-axis to a log (base 2) scale.

```{r}
daydollars %>% ggplot(aes(x = dollars_per_day, y = ..count..)) +
  geom_density() +
  scale_x_continuous(trans = "log2")
```

Create the dollars_per_day variable as in Exercise 7, but for African countries in the years 1970 and 2010 this time. Make sure you remove any NA values. Create a smooth density plot of dollars per day for 1970 and 2010 using a log (base 2) scale for the x axis. Use facet_grid to show a different density plot for 1970 and 2010.

```{r}
gapminder %>% 
  mutate(dollars_per_day = gdp/population/365) %>% 
  filter(continent == "Africa" & year %in% c(1970, 2010) & !is.na(dollars_per_day)) %>% 
  ggplot(aes(x = dollars_per_day, y = ..count..)) +
  geom_density() +
  scale_x_continuous(trans = "log2") +
  facet_grid(year ~ .)
```

Now we are going to edit the code from Exercise 9 to show a stacked density plot of each region in Africa. Make sure the densities are smooth by using bw = 0.5. Use the fill and position arguments where appropriate to create the stacked density plot of each region.

```{r}
gapminder %>% 
  mutate(dollars_per_day = gdp/population/365) %>% 
  filter(continent == "Africa" & year %in% c(1970, 2010) & !is.na(dollars_per_day)) %>% 
  ggplot(aes(x = dollars_per_day, y = ..count.., fill = region)) +
  geom_density(alpha = 0.2, bw = 0.5, position = "stack") +
  scale_x_continuous(trans = "log2") +
  facet_grid(year ~ .)
```

We are going to continue looking at patterns in the gapminder dataset by plotting infant mortality rates versus dollars per day for African countries. Generate dollars_per_day using mutate and filter for the year 2010 for African countries. Store the mutated dataset in gapminder_Africa_2010. Make a scatter plot of infant_mortality versus dollars_per_day for countries in the African continent. Use color to denote the different regions of Africa.

```{r}
gapminder_Africa_2010 <- gapminder %>% 
  mutate(dollars_per_day = gdp/population/365) %>%
  filter(continent == "Africa" & year == 2010 & !is.na(dollars_per_day))
gapminder_Africa_2010 %>% ggplot(aes(dollars_per_day,infant_mortality, color = region)) +
  geom_point()
```

Now we are going to transform the x axis of the plot from the previous exercise. Transform the x axis to be in the log (base 2) scale.

```{r}
gapminder_Africa_2010 %>% ggplot(aes(dollars_per_day,infant_mortality, color = region)) +
  geom_point() +
  scale_x_continuous(trans = "log2")
```

In this exercise, we will remake the plot from Exercise 12 with country names instead of points so we can identify which countries are which. Add a geom_text layer to display country names in addition to of points.

```{r}
gapminder_Africa_2010 %>% ggplot(aes(dollars_per_day,infant_mortality, color = region, label = country)) +
  geom_point() +
  scale_x_continuous(trans = "log2") +
  geom_text()
```

Now we are going to look at changes in the infant mortality and dollars per day patterns African countries between 1970 and 2010. Generate dollars_per_day using mutate and filter for the years 1970 and 2010 for African countries.

```{r}
gapminder %>% 
  mutate(dollars_per_day = gdp/population/365) %>% 
  filter(continent == "Africa" & year %in% c(1970, 2010) & !is.na(dollars_per_day) & !is.na(infant_mortality)) %>% 
  ggplot(aes(dollars_per_day, infant_mortality, color = region, label = country)) +
  geom_point() +
  scale_x_continuous(trans = "log2") +
  geom_text() +
  facet_grid(year ~ .)
```

